---
title: "Assignment 4"
author: "Salma Kinawi"
date: "2025-09-24"
output:
  html_document: default
  pdf_document: default
---




```{r}
#download libraries
library(dplyr)
library(readr)
library(ggplot2)
library(e1071)
library(tree)
library(rpart)
library(gbm)
library(randomForest)
```


```{r}
#import data
data_unedited<-read_csv("Cohen_CANCERSEEK_liquid_biopsy_2018_modified.csv")

#inspect the data
head(data_unedited)
str(data_unedited)
names(data_unedited)
summary(data_unedited)
```

```{r}

#turn tumor type into a factor
data_unedited$Tumor_type<-as.factor(data_unedited$Tumor_type)

#convert the protein columns to become numeric 
data_unedited[, 4:42] <- data.frame(lapply(data_unedited[, 4:42], as.numeric))

```

```{r}
#check for missing values
missing_values<-data.frame(
  Variables=names(data_unedited), 
  Missing=colSums(is.na(data_unedited))
)
missing_values
```

```{r}
#apply median imputation to missing values 

#create a function called impute that will swap an input from the current dataset and replace it with the median of that column only if the vlaue is NA, median is calculated excluding NAs
impute <- function(d) {
  d %>% mutate(across(where(is.numeric),
                       ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))
}  

#apply function to dataset
data_imputed<-impute(data_unedited)

# Check if missing values are gone
colSums(is.na(data_imputed))

```

The preprocessing steps applied to the dataset included converting the Tumor_type column into a factor variable, ensuring that all protein concentration columns were stored as numeric values, and addressing missing data through median imputation. Normalization was not performed, as it is unnecessary for Random Forest models. Unlike distance-based algorithms, Random Forests construct decision trees using threshold splits within each feature independently, making them robust to differences in measurement units or scales across variables.

```{r}
#check dimensions
nrow(data_imputed)
ncol(data_imputed)

#Tumor type counts
table(data_imputed$Tumor_type)

#find how much are not normal
non_normal<-sum((data_imputed$Tumor_type !="Normal"))
non_normal


```

The dataset consists of 1804 rows and 42 columns. A summary of the Tumor_type variable shows that there are 800 normal samples, and by applying the sum() function we determined that there are 1004 tumor samples distributed across various cancer types representing different body parts and organs.

```{r}
#make new column sorting into odd or even
data_imputed$split <- ifelse(seq_len(nrow(data_imputed)) %% 2==1, "Train", "Test")

#view new column, split 
head(data_imputed)

#split into two datasets

train<-subset(data_imputed, split=="Train")

test<-subset(data_imputed, split=="Test")

#remove non predictor columns
train<- subset(train, select = -c(Patient_ID, Sample_ID,split))
test <- subset(test,  select = -c(Patient_ID, Sample_ID, split))


head(train)
head(test)
```

```{r}
# Make decision tree model
treemodel <- tree(Tumor_type ~ ., data = train)
summary(treemodel)
plot(treemodel)
text(treemodel, cex=0.7, adj=0)
```
```{r}
#find the most influential protein feature
head(treemodel$frame[, c("var","n","dev","yval")]) #looking at what the frame of the tree model looks like

#find the most influential protein
influential<- treemodel$frame$var[1]
influential
```

```{r}
#predictions using the test dataset 
predictions<- predict(treemodel, test, type = "class")

#make a confusion matrix
matrix<-table(predicted=predictions, actual=test$Tumor_type)
matrix 

#find the misclassifcation rate
misclass<-mean(predictions !=test$Tumor_type)
misclass


```
This analysis shows that the decision tree had a misclassification error rate of 0.3725, meaning that approximately 37.3% of the test samples were misclassified. The model performed well in predicting Normal, Colorectum, Breast, and Pancreas samples. However, it struggled with several cancer types: Esophagus and Liver were never correctly classified, while Ovary and Lung showed relatively poor prediction accuracy, with many samples misclassified into other categories.

Part 3:


```{r}
TrainModel<-train
TestModel<-test

names(TrainModel) <- make.names(names(TrainModel), unique = TRUE)
names(TestModel)  <- make.names(names(TestModel),  unique = TRUE)

#make results reproducible
set.seed(53)

#number of protein predictors minus the response Tumor_type
protein<-ncol(train)-1 

randomf_model <- randomForest(Tumor_type ~ .,data = train,ntree = 500,
 mtry  = floor(sqrt(protein)),   # good default; you had 6, either is fine
  importance = TRUE
)
randomf_model 
```
```{r}
#predit class lables on the test set
rf_pred<-predict(randomf_model, newdata = test)

#create confusion matrix
conf_matrix<-table(predicted=rf_pred, actual=test$Tumor_type)
conf_matrix

#miclassification error rate
m_error <-mean(rf_pred !=test$Tumor_type)
m_error
```

```{r}
# Variable importance values
imp<-importance(randomf_model)
imp
#importance plot
df_imp <- data.frame(
  Feature = rownames(imp),
  Score   = imp[, "MeanDecreaseAccuracy"]
)
#order by score
df_imp <- df_imp[order(df_imp$Score), ]
tail(df_imp, 10)


#use the top 10 features since it is really crowded
df_subset<- tail(df_imp, 10)

#make barplot
barplot(
  df_subset$Score,
  names.arg = df_subset$Feature,
  las = 2,                      
  cex.names = 0.7, 
  col = "grey",
        main = paste("Top 10 Protein Importances"),
        ylab = "Mean Decrease Accuracy")

```
From the importance table, IL-8 has the highest Mean Decrease in Accuracy (28.98), making it the most influential protein for classifying tumor type.

```{r}
#cancer vs normal
train$binary <- ifelse(train$Tumor_type =="Normal", "Normal", "Cancer")
test$binary <-ifelse(test$Tumor_type=="Normal", "Normal", "Cancer")

#turn into a factor
train$binary <- factor(train$binary, levels = c("Normal","Cancer"))
test$binary  <- factor(test$binary,  levels = c("Normal","Cancer"))

table(train$binary); table(test$binary)

```
```{r}
set.seed(443)
pro<-ncol(train) - 2

rf_binary <- randomForest(
  binary ~ . - Tumor_type,
  data = train,
  ntree = 500,
  mtry  = floor(sqrt(pro)),
  importance = TRUE
)
rf_binary
```
```{r}
#predict on test dataset
pred<-predict(rf_binary, newdata = test, type="class")
#confusions matrix
cof_matrix<- table(predicted = pred, 
                   actual=test$binary)
cof_matrix

#missclassification rate
missclass<-mean(pred !=test$binary)
accuracy<-1-missclass #overall accuracy

accuracy
missclass

```
```{r}
#cohen's Kappa
total<- sum(cof_matrix)
pe<-sum(rowSums(cof_matrix) * colSums(cof_matrix)) / (total^2)
kappa <- (accuracy - pe) / (1 - pe)
kappa

```

```{r}
#importance
imp_binary<-importance(rf_binary)

#convert into a data frame
df_imp_binary <- data.frame(
  feature = rownames(imp_binary),
  score   = imp_binary[, "MeanDecreaseAccuracy"],
  row.names = NULL
)

# Sort by importance
df_imp_binary <- df_imp_binary[order(df_imp_binary$score), ]
df_subset <- tail(df_imp_binary, 10) # Take top 10


scores   <- as.numeric(df_subset$score)
features <- as.character(df_subset$feature)

barplot(
  height = scores,
  names.arg = features,
  las = 2,
  cex.names = 0.7,
  col = "purple",
  main = "Top 10 Protein Importances (Binary)",
  ylab = "Mean Decrease Accuracy"
)

#top two 
df_imp_binary$feature[order(df_imp_binary$score, decreasing = TRUE)[1:2]]
```
3. The two most influential proteins identified were IL-8 and IL-6.

4. A biological explanation for this finding is that IL-8 is a pro-inflammatory chemokine that promotes the formation of new blood vessels (angiogenesis) and is often elevated in cancer patients, supporting tumor growth and metastasis. IL-6 is a cytokine with key roles in inflammation and immune regulation, and it is commonly associated with cancer-related inflammation and tumor progression. Together, these markers explain why IL-8 and IL-6 strongly discriminate between cancer and normal blood samples.

5. In terms of model usefulness, the binary random forest achieved a misclassification rate of 5.7%, an accuracy of 94%, and a Cohen’s kappa of 0.88, indicating strong agreement beyond chance. These results demonstrate that the binary model performs very well in distinguishing between cancer and normal samples, suggesting its potential value for detecting cancer in blood samples using this protein panel.
```